New cardinality estimation algorithms for HyperLogLog sketches

### 2016-11-27 paper/paper.tex
added appendix that derives the deviation of function xi from constant 1

% SYMBOLS
\newcommand*{\symDefine}[2]{\newcommand{{#1}}{{#2}}}
\symDefine{\symPrecision}{p}
\symDefine{\symRegRange}{q}
\symDefine{\symRegVal}{k}
\symDefine{\symRegValVariate}{K}
\symDefine{\symNumReg}{m}
\symDefine{\symDataItem}{D}
\symDefine{\symBitRepA}{a}
\symDefine{\symBitRepB}{b}
\symDefine{\symIndexI}{i}
\symDefine{\symIndexJ}{j}
\symDefine{\symIndexL}{l}
\symDefine{\symCount}{c}
\symDefine{\symCountVariate}{C}
\symDefine{\symAlpha}{\alpha}
\symDefine{\symBeta}{\beta}
\symDefine{\symA}{a}
\symDefine{\symB}{b}
\symDefine{\symKappa}{\kappa}
\symDefine{\symX}{x}
\symDefine{\symY}{y}
\symDefine{\symZ}{z}
\symDefine{\symPhi}{\varphi}
\symDefine{\symXEstimate}{\hat{\symX}}
\symDefine{\symXNormalized}{\symX'}
\symDefine{\symN}{n}
\symDefine{\symS}{s}
\symDefine{\symInteger}{z}
\symDefine{\symIntegerVariate}{Z}
\symDefine{\symCardinality}{n}
\symDefine{\symCardinalityEstimate}{\hat{\symCardinality}}
\symDefine{\symCardinalityRawEstimate}{\symCardinalityEstimate_\text{raw}}
\symDefine{\symError}{\varepsilon}
\symDefine{\symStopDelta}{\delta}
\symDefine{\symStopEpsilon}{\varepsilon}
\symDefine{\symBigO}{\mathcal{O}}
\symDefine{\symExpectation}{\mathbb{E}}
\symDefine{\symProbability}{P}
\symDefine{\symProbabilityMass}{\rho}
\symDefine{\symRegProbability}{\gamma}
\symDefine{\symLikelihood}{\mathcal{L}}
\symDefine{\symPoissonRate}{\lambda}
\symDefine{\symPoissonRateEstimate}{\hat{\symPoissonRate}}
\symDefine{\symFunc}{f}
\symDefine{\symFuncPrime}{g}
\symDefine{\symHelper}{h}
\symDefine{\symHelperApprox}{\tilde{\symHelper}}
\symDefine{\symHelperW}{w}
\symDefine{\symHelperV}{v}
\symDefine{\symHelperU}{u}
\symDefine{\symSetA}{A}
\symDefine{\symSetB}{B}
\symDefine{\symSetS}{S}
\symDefine{\symSetX}{X}
\symDefine{\symSetASuffix}{a}
\symDefine{\symSetBSuffix}{b}
\symDefine{\symSetXSuffix}{x}
\symDefine{\symCountMatrix}{\mathbf{c}}
\symDefine{\symPowerSeriesFunc}{\xi}
\symDefine{\symSmallCorrectionFunc}{\sigma}
\symDefine{\symLargeCorrectionFunc}{\tau}
\symDefine{\symEpsPowerSeriesFunc}{\varepsilon_\symPowerSeriesFunc}
\symDefine{\symBiasCorrectionFunc}{w_{\text{corr}}}

@@

Let us assume that the register values have following cumulative distribution function
\begin{equation}
\label{equ:assumed_register_val_distribution}
\symProbability(\symRegValVariate \leq \symRegVal\vert\symPoissonRate) = e^{-\frac{\symPoissonRate}{\symNumReg 2^{ \symRegVal}}}.
\end{equation}
For now we ignore that this distribution has infinite support and differs from the register value distribution under the Poisson model \eqref{equ:register_value_distribution}, whose support is limited to the range $[0, \symRegRange+1]$. For a random variable $\symRegValVariate$ obeying \eqref{equ:assumed_register_val_distribution} the expectation of $2^{-\symRegValVariate}$ is given by
\begin{multline}
\label{equ:expectation_power_two}
\symExpectation(2^{-\symRegValVariate})
=
\\
\sum_{\symRegVal = -\infty}^\infty
2^{-\symRegVal}
\left(
e^{-\frac{\symPoissonRate}{\symNumReg 2^{\symRegVal}}}
-e^{-\frac{\symPoissonRate}{\symNumReg 2^{\symRegVal-1}}}
\right)
=
\frac{1}{2}
\sum_{\symRegVal = -\infty}^\infty
2^{\symRegVal}
e^{-\frac{\symPoissonRate}{\symNumReg} 2^{\symRegVal}}
=
\frac{\symAlpha_\infty\,\symNumReg\,\symPowerSeriesFunc\!\left(\log_2\!\left(\symPoissonRate/\symNumReg\right)\right)}{\symPoissonRate},
\end{multline}
where the function 
\begin{equation}
\label{equ:power_series_function}
\symPowerSeriesFunc(\symX):= \log(2) \sum_{\symRegVal = -\infty}^\infty
2^{\symRegVal+\symX}
e^{-2^{\symRegVal+\symX}}
\end{equation}
is a smooth periodic function with period 1. 
%Its Fourier series is
%\begin{equation}
%\symPowerSeriesFunc(\symX) = 1 + \sum_{\symIndexL=1}^{\infty}
%\symA_\symIndexL \cos(2\pi\symIndexL\symX) + \symB_\symIndexL \sin(2\pi\symIndexL\symX)
%\end{equation}
%with coefficients
%\begin{equation}
%\begin{aligned}
%\symA_\symIndexL &= 2\log(2)\int_{-\infty}^\infty 2^\symX e^{-2^\symX}\cos(2\pi\symIndexL\symX)d\symX, \\
%\symB_\symIndexL &= 2\log(2)\int_{-\infty}^\infty 2^\symX e^{-2^\symX}\sin(2\pi\symIndexL\symX)d\symX.
%\end{aligned}
%\end{equation}
%Numerical evaluations give for the $\symA_1\approx \num{6.353e-6}$ and $\symB_1\approx \num{-7.572e-6}$.
Numerical evaluations indicate that this function can be bounded by $1 - \symEpsPowerSeriesFunc 
\leq \symPowerSeriesFunc(\symX) \leq 1 + \symEpsPowerSeriesFunc$ with $\symEpsPowerSeriesFunc:=\num{9.885e-6}$ (see \cref{fig:power_series_func}).

@@

The more relevant case with stochastic averaging, which is considered in this paper, is more complicated. However, with the help of the Poisson approximation, we are able to still apply the maximum likelihood method and to derive a new robust and efficient cardinality estimation algorithm. Furthermore, we will demonstrate that consequent application of the maximum likelihood method reveals that the cardinality estimate needs to be roughly proportional to the harmonic mean for intermediate cardinality values. The history of the HyperLogLog algorithm shows that the raw estimator \eqref{equ:raw_estimator} was first found after several attempts using the geometric mean \cite{Flajolet2007, Durand2003}.

@@

The HyperLogLog algorithm itself can be regarded as hashing algorithm as it maps sets to register values. For sufficiently large cardinalities we can use the Poisson approximation and assume that the number of zero-valued HyperLogLog register can be ignored. Furthermore, if the HyperLogLog parameter $\symRegRange$ is chosen large enough, the number of saturated registers can be ignored as well. As a consequence, we can assume that the distribution of register values follows \eqref{equ:assumed_register_val_distribution} and the probability that a register has the same value for two different sets is (compare \eqref{equ:register_value_joint_pmf})
\begin{equation}
\symProbability(\symRegValVariate_1=\symRegValVariate_2)
=
\sum_{\symRegVal = -\infty}^{\infty}
e^{-\frac{\symPoissonRate_\symSetASuffix + \symPoissonRate_\symSetBSuffix + \symPoissonRate_\symSetXSuffix}
{\symNumReg 2^\symRegVal}
}
\left(
1
-
e^{-\frac{\symPoissonRate_\symSetASuffix +  \symPoissonRate_\symSetXSuffix}
{\symNumReg 2^\symRegVal}
}
-
e^{-\frac{\symPoissonRate_\symSetBSuffix + \symPoissonRate_\symSetXSuffix}
{\symNumReg 2^\symRegVal}
}
+
e^{-\frac{\symPoissonRate_\symSetASuffix + \symPoissonRate_\symSetBSuffix + \symPoissonRate_\symSetXSuffix}
{\symNumReg 2^\symRegVal}
}
\right).
\end{equation}
Using the approximation 
$\sum_{\symRegVal=-\infty}^\infty
e^{-\frac{\symX}{2^\symRegVal}} - e^{-\frac{\symY}{2^\symRegVal}}
\approx
2\symAlpha_\infty \left(\log(\symY)-\log(\symX)\right)$ we get
\begin{equation}
\symProbability(\symRegValVariate_1=\symRegValVariate_2)
\approx
1
+
2\symAlpha_\infty
\left(
\log\left(
1
-
\frac{1}{2}
\frac{\symPoissonRate_\symSetASuffix}
{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+
\symPoissonRate_\symSetXSuffix}
\right)
+
\log\left(
1
-
\frac{1}{2}
\frac{\symPoissonRate_\symSetBSuffix}
{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+
\symPoissonRate_\symSetXSuffix}
\right)
\right).
\end{equation}
Further reduction by using the first order approximation $\log(1-\symX)\approx -\symX$ finally shows that the probability of a register having the same value for both sets is roughly linear to the Jaccard index $\frac{\symPoissonRate_\symSetXSuffix}
{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+
\symPoissonRate_\symSetXSuffix}$
\begin{equation}
\symProbability(\symRegValVariate_1=\symRegValVariate_2)
\approx
\symAlpha_\infty
\frac{\symPoissonRate_\symSetXSuffix}
{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+
\symPoissonRate_\symSetXSuffix}
+
(1-\symAlpha_\infty).
\end{equation}
This finding makes the HyperLogLog algorithm an interesting candidate for locality-sensitive hashing. What also looks promising is the much simpler preprocessing because it only involves a single hash function (compare \cref{alg:insert}). In contrast, min-wise hashing is very costly, because many hash function evaluations are necessary for all the needed permutations \cite{Li2011}.

@@

>>>>

% SYMBOLS
\renewcommand\Re{\operatorname{Re}}
\renewcommand\Im{\operatorname{Im}}

\newcommand*{\symDefine}[2]{\newcommand{{#1}}{{#2}}}
\symDefine{\symPrecision}{p}
\symDefine{\symRegRange}{q}
\symDefine{\symRegVal}{k}
\symDefine{\symRegValVariate}{K}
\symDefine{\symNumReg}{m}
\symDefine{\symDataItem}{D}
\symDefine{\symBitRepA}{a}
\symDefine{\symBitRepB}{b}
\symDefine{\symIndexI}{i}
\symDefine{\symImaginary}{\rm{i}}
\symDefine{\symIndexJ}{j}
\symDefine{\symIndexL}{l}
\symDefine{\symCount}{c}
\symDefine{\symCountVariate}{C}
\symDefine{\symAlpha}{\alpha}
\symDefine{\symBeta}{\beta}
\symDefine{\symA}{a}
\symDefine{\symB}{b}
\symDefine{\symKappa}{\kappa}
\symDefine{\symX}{x}
\symDefine{\symY}{y}
\symDefine{\symZ}{z}
\symDefine{\symPhi}{\varphi}
\symDefine{\symXEstimate}{\hat{\symX}}
\symDefine{\symXNormalized}{\symX'}
\symDefine{\symN}{n}
\symDefine{\symS}{s}
\symDefine{\symInteger}{z}
\symDefine{\symIntegerVariate}{Z}
\symDefine{\symCardinality}{n}
\symDefine{\symCardinalityEstimate}{\hat{\symCardinality}}
\symDefine{\symCardinalityRawEstimate}{\symCardinalityEstimate_\text{raw}}
\symDefine{\symError}{\varepsilon}
\symDefine{\symStopDelta}{\delta}
\symDefine{\symStopEpsilon}{\varepsilon}
\symDefine{\symBigO}{\mathcal{O}}
\symDefine{\symExpectation}{\mathbb{E}}
\symDefine{\symProbability}{P}
\symDefine{\symProbabilityMass}{\rho}
\symDefine{\symRegProbability}{\gamma}
\symDefine{\symLikelihood}{\mathcal{L}}
\symDefine{\symPoissonRate}{\lambda}
\symDefine{\symPoissonRateEstimate}{\hat{\symPoissonRate}}
\symDefine{\symFunc}{f}
\symDefine{\symFuncPrime}{g}
\symDefine{\symHelper}{h}
\symDefine{\symHelperApprox}{\tilde{\symHelper}}
\symDefine{\symHelperW}{w}
\symDefine{\symHelperV}{v}
\symDefine{\symHelperU}{u}
\symDefine{\symSetA}{A}
\symDefine{\symSetB}{B}
\symDefine{\symSetS}{S}
\symDefine{\symSetX}{X}
\symDefine{\symSetASuffix}{a}
\symDefine{\symSetBSuffix}{b}
\symDefine{\symSetXSuffix}{x}
\symDefine{\symCountMatrix}{\mathbf{c}}
\symDefine{\symPowerSeriesFunc}{\xi}
\symDefine{\symSmallCorrectionFunc}{\sigma}
\symDefine{\symLargeCorrectionFunc}{\tau}
\symDefine{\symEpsPowerSeriesFunc}{\varepsilon_\symPowerSeriesFunc}
\symDefine{\symBiasCorrectionFunc}{w_{\text{corr}}}

@@

Let us assume that the register values have following cumulative distribution function
\begin{equation}
\label{equ:assumed_register_val_distribution}
\symProbability(\symRegValVariate \leq \symRegVal\vert\symPoissonRate) = e^{-\frac{\symPoissonRate}{\symNumReg 2^{ \symRegVal}}}.
\end{equation}
For now we ignore that this distribution has infinite support and differs from the register value distribution under the Poisson model \eqref{equ:register_value_distribution}, whose support is limited to the range $[0, \symRegRange+1]$. For a random variable $\symRegValVariate$ obeying \eqref{equ:assumed_register_val_distribution} the expectation of $2^{-\symRegValVariate}$ is given by
\begin{multline}
\label{equ:expectation_power_two}
\symExpectation(2^{-\symRegValVariate})
=
\\
\sum_{\symRegVal = -\infty}^\infty
2^{-\symRegVal}
\left(
e^{-\frac{\symPoissonRate}{\symNumReg 2^{\symRegVal}}}
-e^{-\frac{\symPoissonRate}{\symNumReg 2^{\symRegVal-1}}}
\right)
=
\frac{1}{2}
\sum_{\symRegVal = -\infty}^\infty
2^{\symRegVal}
e^{-\frac{\symPoissonRate}{\symNumReg} 2^{\symRegVal}}
=
\frac{\symAlpha_\infty\,\symNumReg\,\symPowerSeriesFunc\!\left(\log_2\!\left(\symPoissonRate/\symNumReg\right)\right)}{\symPoissonRate},
\end{multline}
where the function 
\begin{equation}
\label{equ:power_series_function}
\symPowerSeriesFunc(\symX):= \log(2) \sum_{\symRegVal = -\infty}^\infty
2^{\symRegVal+\symX}
e^{-2^{\symRegVal+\symX}}
\end{equation}
is a smooth periodic function with period 1. 
Numerical evaluations indicate that this function can be bounded by $1 - \symEpsPowerSeriesFunc 
\leq \symPowerSeriesFunc(\symX) \leq 1 + \symEpsPowerSeriesFunc$ with $\symEpsPowerSeriesFunc:=\num{9.885e-6}$ (see \cref{fig:power_series_func}). This value can also be found using Fourier analysis as shown in in \cref{app:analysis_xi}.

@@

The more relevant case with stochastic averaging, which is considered in this paper, is more complicated. However, with the help of the Poisson approximation, it is still feasible to apply the maximum likelihood method and we are able to derive a new robust and efficient cardinality estimation algorithm. Furthermore, we will demonstrate that consequent application of the maximum likelihood method reveals that the cardinality estimate needs to be roughly proportional to the harmonic mean for intermediate cardinality values. The history of the HyperLogLog algorithm shows that the raw estimator \eqref{equ:raw_estimator} was first found after several attempts using the geometric mean \cite{Flajolet2007, Durand2003}.

@@

The HyperLogLog algorithm itself can be regarded as hashing algorithm as it maps sets to register values. For sufficiently large cardinalities we can use the Poisson approximation and assume that the number of zero-valued HyperLogLog register can be ignored. Furthermore, if the HyperLogLog parameter $\symRegRange$ is chosen large enough, the number of saturated registers can be ignored as well. As a consequence, we can assume that the distribution of register values follows \eqref{equ:assumed_register_val_distribution} and the probability that a register has the same value for two different sets is (compare \eqref{equ:register_value_joint_pmf})
\begin{equation}
\symProbability(\symRegValVariate_1=\symRegValVariate_2)
=
\sum_{\symRegVal = -\infty}^{\infty}
e^{-\frac{\symPoissonRate_\symSetASuffix + \symPoissonRate_\symSetBSuffix + \symPoissonRate_\symSetXSuffix}
{\symNumReg 2^\symRegVal}
}
\left(
1
-
e^{-\frac{\symPoissonRate_\symSetASuffix +  \symPoissonRate_\symSetXSuffix}
{\symNumReg 2^\symRegVal}
}
-
e^{-\frac{\symPoissonRate_\symSetBSuffix + \symPoissonRate_\symSetXSuffix}
{\symNumReg 2^\symRegVal}
}
+
e^{-\frac{\symPoissonRate_\symSetASuffix + \symPoissonRate_\symSetBSuffix + \symPoissonRate_\symSetXSuffix}
{\symNumReg 2^\symRegVal}
}
\right).
\end{equation}
Using the approximation 
$\sum_{\symRegVal=-\infty}^\infty
e^{-\frac{\symX}{2^\symRegVal}} - e^{-\frac{\symY}{2^\symRegVal}}
\approx
2\symAlpha_\infty \left(\log(\symY)-\log(\symX)\right)$ we get
\begin{equation}
\symProbability(\symRegValVariate_1=\symRegValVariate_2)
\approx
1
+
2\symAlpha_\infty
\left(
\log\!\left(
1
-
\frac{1}{2}
\frac{\symPoissonRate_\symSetASuffix}
{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+
\symPoissonRate_\symSetXSuffix}
\right)
+
\log\!\left(
1
-
\frac{1}{2}
\frac{\symPoissonRate_\symSetBSuffix}
{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+
\symPoissonRate_\symSetXSuffix}
\right)
\right).
\end{equation}
Further reduction by using the first order approximation $\log(1-\symX)\approx -\symX$ finally shows that the probability of a register having the same value for both sets is roughly linear to the Jaccard index $\frac{\symPoissonRate_\symSetXSuffix}
{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+
\symPoissonRate_\symSetXSuffix}$
\begin{equation}
\symProbability(\symRegValVariate_1=\symRegValVariate_2)
\approx
\symAlpha_\infty
\frac{\symPoissonRate_\symSetXSuffix}
{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+
\symPoissonRate_\symSetXSuffix}
+
(1-\symAlpha_\infty).
\end{equation}
This finding makes the HyperLogLog algorithm an interesting candidate for locality-sensitive hashing. What also looks promising is the much simpler preprocessing because it only involves a single hash function (compare \cref{alg:insert}). In contrast, min-wise hashing is very costly, because many hash function evaluations are necessary for all the needed permutations \cite{Li2011}.

@@

\section{Analysis of $\symPowerSeriesFunc(\symX)$}
\label{app:analysis_xi}
The Fourier series of the function $\symPowerSeriesFunc(\symX)$, which is defined by
\begin{equation}
\symPowerSeriesFunc(\symX):= \log(2) \sum_{\symRegVal = -\infty}^\infty
2^{\symRegVal+\symX}
e^{-2^{\symRegVal+\symX}}
\end{equation}
and which has period equal to 1, is
\begin{equation}
\symPowerSeriesFunc(\symX) = 1 + \Re\!\left(\sum_{\symIndexL=1}^{\infty}
\symA_\symIndexL \exp(2\pi\symIndexL\symX\symImaginary)\right)
\end{equation}
with coefficients
\begin{equation}
\symA_\symIndexL = 2\int_0^1 \symPowerSeriesFunc(\symX) \exp(-2\pi\symIndexL\symX\symImaginary)d\symX
=
2\log(2) \int_{-\infty}^{\infty} 2^\symX
e^{-2^\symX}\exp(-2\pi\symIndexL\symX\symImaginary)d\symX.
\end{equation}
The variable transformation $\symY = 2^\symX$ yields
\begin{equation}
\symA_\symIndexL
=
2\int_{0}^{\infty} e^{-\symY}
\symY^{-\frac{2\pi\symIndexL\symImaginary}{\log(2)}}d\symY
= 
2\Gamma\!\left(1-\frac{2\pi\symIndexL\symImaginary}{\log(2)}\right)
\end{equation}
where $\Gamma$ denotes the gamma function. Using the identity $\left|\Gamma(1+\symImaginary\symX)\right| = 
\sqrt{\frac{\pi\symX}{\sinh(\pi\symX)}}
$
we are able to write for the absolute values of the coefficients
\begin{equation}
|\symA_\symIndexL| = 
2\sqrt{\frac{\symB\symIndexL}{\sinh(\symB\symIndexL)}}
\quad
\text{with}
\
\symB:=\frac{2\pi^2}{\log 2}.
\end{equation}
In particular, the amplitude of the first harmonic is 
$|\symA_1|
\approx
\num{9.884e-6}$. Obviously, the maximum deviation of $\symPowerSeriesFunc(\symX)-1$ from the first harmonic must be smaller than $\sum_{\symIndexL=2}^{\infty} 
|\symA_\symIndexL|$. The ratio of subsequent coefficients is given by
\begin{equation}
\frac{|\symA_{\symIndexL+1}|}{|\symA_{\symIndexL}|}
=
\sqrt{\frac{\symIndexL+1}{\symIndexL}}
\sqrt{\frac{
\sinh(\symB \symIndexL)
}{
\sinh(\symB (\symIndexL+1))
}}
=
\sqrt{\frac{\symIndexL+1}{\symIndexL}}
\sqrt{\frac{
1
}{
\cosh(\symB)
+
\frac{
\sinh(\symB)
}
{
\tanh(\symB \symIndexL)
}
}}.
\end{equation}
For $\symIndexL\geq 2$ we have $\sqrt{\frac{\symIndexL+1}{\symIndexL}}\leq \sqrt{\frac{3}{2}}$. Together with $\tanh(x)\leq 1$ we obtain
\begin{equation}
\frac{|\symA_{\symIndexL+1}|}{
|\symA_{\symIndexL}|
}
\leq
\sqrt{\frac{3}{2}}
\sqrt{\frac{
1
}{
\cosh(\symB)
+
\sinh(\symB)
}}
\\
=
\sqrt{\frac{
3
}{
2 e^\symB
}}.
\end{equation}
Consequently, $|\symA_{\symIndexL}| \leq 
|\symA_{2}|
\left(\sqrt{\frac{
3
}{
2 e^\symB
}}\right)^{\symIndexL-2}$ for $\symIndexL\geq 2$ and 
\begin{equation}
\sum_{\symIndexL=2}^{\infty} 
|\symA_\symIndexL|
\leq
|\symA_2|
\sum_{\symIndexL=0}^{\infty} 
\left(
\sqrt{\frac{
3
}{
2e^\symB
}}
\right)^\symIndexL
=
2\sqrt{
\frac{2\symB}{\sinh(2\symB)}}
\frac{1}{1-
\sqrt{\frac{
3
}{
2 e^\symB
}}
}
\approx
\num{9.154e-12}.
\end{equation}
For the maximum deviation from 1 we finally have
\begin{equation}
\max(|\symPowerSeriesFunc(\symX)-1|) \leq \sum_{\symIndexL=1}^{\infty} 
|\symA_\symIndexL|
\leq \num{9.885e6}.
\end{equation}
